### 2025-03-10 paper/sections/abstract.qmd

Counterfactual Explanations have emerged as a popular tool to explain predictions made by opaque machine learning models: they explain how factual inputs need to change in order for some fitted model to produce some desired output. Much existing research has focused on identifying explanations that are not only valid but also deemed desirable with respect to the underlying data and stakeholder requirements. Recent work has shown that under this premise, the task of learning desirable explanations is effectively reassigned from the model itself to the (post-hoc) counterfactual explainer. Building on that work, we propose a novel model objective that leverages counterfactuals during the training phase (ad-hoc) in order to minimize the divergence between learned representations and desirable explanations. Through extensive experiments, we demonstrate that our proposed methodology facilitates training models that inherently deliver desirable explanations while maintaining high predictive performance.

>>>>

We propose a novel training regime called Counterfactual Training that leverages counterfactual explanations to increase the explanatory capacity of models. Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output. To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to stakeholder requirements. Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata. In this work, we instead hold models directly accountable for this desired end goal: Counterfactual Training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations. We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while maintaining high predictive performance.

### 2025-03-10 paper/sections/intro.qmd

Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern AIs are tasked with learning these representations from scratch, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. Modern advances in computing have made it possible to provide such AIs with ever greater degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models. Unfortunately, in doing so they also learn increasingly complex and highly sensitive representations that we can no longer easily interpret.

This trend towards complexity for the sake of performance has come under serious scrutiny in recent years. At the very cusp of the deep learning revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are sensitive to adversarial examples: counterfactuals of model inputs that yield vastly different model predictions despite being "imperceptible" in that they are semantically indifferent from their factual counterparts. Despite partially effective mitigation strategies such as **adversarial training** [@goodfellow2014explaining], truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].

Part of the problem is that high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case]. Based purely on predictive performance, these solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns. This poses two related challenges: firstly, it makes these models inherently opaque, since humans cannot simply interpret what type of explanation the complex learned representations correspond to; secondly, even if we could resolve the first challenge, it is not obvious how to mitigate models from learning representations that correspond to meaningless and undesirable explanations.

The first challenge has attracted an abundance of research on **explainable AI** (XAI) which aims to develop tools to derive explanations from complex model representations. This can mitigate a scenario in which we deploy opaque models and blindly rely on their predictions. On countless occasions, this scenario has already occurred in practice and caused real harm to people who were affected adversely and often unfairly by automated decision-making systems (ADMS) involving opaque models [@oneil2016weapons]. Effective XAI tools can aide us in monitoring models and providing recourse to individuals to turn adverse outcomes (e.g. "loan application rejected") into positive ones ("application accepted"). @wachter2017counterfactual propose **counterfactual explanations** as an effective approach to achieve this: they explain how factual inputs need to change in order for some fitted model to produce some desired output, typically involving minimal perturbations.

To our surprise, the second challenge has not yet attracted any consolidated research effort. Specifically, there has been no concerted effort towards improving model **explainability**, which we define here as the degree to which learned representations correspond to explanations that are interpretable and deemed desirable by humans. Instead, the choice has typically been to improve the capacity of XAI tools to identify the subset explanations that are both desirable and valid for any given model, independent of whether the learned representations are also compatible with undesirable explanations [@altmeyer2024faithful]. Fortunately, recent findings indicate that explainability can arise as byproduct of regularization techniques aimed at other objectives such as robustness, generalization and generative capacity [@schut2021generating, @augustin2020adversarial, @altmeyer2024faithful].

Building on these findings, we introduce **counterfactual training**: a novel regularization technique geared explicitly towards aligning model representations with desirable explanations. Our contributions are as follows:

@@

Despite limitations of our approach discussed in @sec-discussion, we conclude that counterfactual training provides a practical framework for researchers and practitioners interested in making opaque models more trustworthy [@sec-conclusion]. We also believe that this work serves as an opportunity for XAI researchers to reevaluate the premise of improving XAI tools without improving models.

>>>>

Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern machine learning (ML) models are tasked with learning the representations directly from data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. Modern advances in computing have made it possible to provide such models with ever growing degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models. Unfortunately, in doing so, the models learn increasingly complex and highly sensitive representations that we can no longer easily interpret.

The trend towards complexity for the sake of performance has come under serious scrutiny in recent years. At the very cusp of the deep learning revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are sensitive to adversarial examples: counterfactuals of model inputs that yield vastly different model predictions despite being "imperceptible" in that they are semantically indifferent from their factual counterparts. Although some partially effective mitigation strategies have been proposed, for example **adversarial training** [@goodfellow2014explaining], truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].

Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so called "lottery tickets" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery] and generalizability [@morcos2019success]. Similarly, @zhang2021understanding showed that state-of-the-art neural networks are so expressive that they can fit randomly labeled data. Thus, looking at the predictive performance, the solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns. This poses two related challenges. Firstly, there is no dependable way to verify if such complex representations correspond to meaningful and plausible explanations. Secondly, even if we could resolve the first challenge, it remains undecided how to ensure that models can *only* learn valuable explanations.

The first challenge has attracted an abundance of research on **explainable AI** (XAI) which aims to develop tools to derive explanations from complex model representations. This can mitigate a scenario in which we deploy opaque models and blindly rely on their predictions. On countless occasions, this scenario has occurred in practice and caused real harm to people who were affected adversely and often unfairly by automated decision-making (ADM) systems involving opaque models [@oneil2016weapons; @mcgregor2021preventing]. Effective XAI tools can aid us in monitoring models and providing recourse to individuals to turn adverse outcomes (e.g., "loan application rejected") into positive ones (e.g., "application accepted"). @wachter2017counterfactual propose **counterfactual explanations** (CE) as an effective approach to achieve this goal: CEs explain how factual inputs need to change in order for some fitted model to produce some desired output, typically involving minimal perturbations.

To our surprise, the second challenge has not yet attracted any major consolidated research effort. Specifically, there has been no concerted effort towards improving improving models' explanatory capacity, which we will henceforth simply call "explainability", defined as the degree to which learned representations correspond to explanations that are interpretable and deemed **plausible** by humans (see @def-explainability). Instead, the choice has typically been to improve the ability of XAI tools to identify the subset explanations that are both plausible and valid for any given model, independent of whether the learned representations are also compatible with implausible explanations [@altmeyer2024faithful]. Fortunately, recent findings indicate that explainability can arise as byproduct of regularization techniques aimed at other objectives such as robustness, generalization, and generative capacity [@schut2021generating; @augustin2020adversarial; @altmeyer2024faithful].

Building on these findings, we introduce **counterfactual training**: a novel training regime explicitly geared towards aligning model representations with plausible explanations. Our contributions are as follows:

@@

Despite some limitations of our approach discussed in @sec-discussion, we conclude in @sec-conclusion that counterfactual training provides a practical framework for researchers and practitioners interested in making opaque models more trustworthy. We also believe that this work serves as an opportunity for XAI researchers to re-evaluate the trend of improving XAI tools without improving the underlying models.

### 2025-03-10 paper/sections/method.qmd

Counterfactual training combines ideas from adversarial training, energy-based modelling and counterfactuals explanations with the explicit objective of aligning representations with plausible explanations that comply with user requirements. In the context of CE, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful]. Plausibility is a necessary but insufficient condition for using CE to provide algorithmic recourse (AR) to individuals affected by opaque models in practice. This is because for recourse recommendations to be **actionable**, they need to not only result in plausible counterfactuals but also be attainable. A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they were 20 years older. Ignoring all other features, this complies with the definition of plausibility if 40-year-old individuals are in fact more credit-worthy on average than young adults. But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a mutable feature. For our intents and purposes, counterfactual training aims at improving model explainability by aligning models with counterfactuals that meet both desiderata, plausibility and actionability. Formally, we define explainability as follows:

@@

1. (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
2. (Actionability) Permutations $\Delta$ are subject to actionability constraints.

@@

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual explanation generated through gradient descent over $T$ iterations as initially proposed by @wachter2017counterfactual. For our purposes, we let $T$ vary and consider the counterfactual search as converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$^[For detailed background information on gradient-based counterfactual search and convergence see @sec-app-ce.].

To train models with high explainability as defined in @def-explainability, we propose to leverage counterfactuals in the following objective,

@@

where $\text{yloss}(\cdot)$ is any conventional classification loss that induces discriminative performance (e.g. cross-entropy). The two additional components in @eq-obj are explained in more detail below. For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: 1) the contrastive divergence, $\text{div}(\cdot)$, between counterfactuals $\mathbf{x}_T^\prime$ and observed samples $x$ and, 2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. The tradeoff between the different components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$.

@@

@grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on inputs and generate inputs. They show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$. The first factor can be optimized using conventional cross-entropy as in @eq-obj. To optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimize the contrastive divergence between samples drawn from $p_\theta(\mathbf{x})$ and training observations, i.e. samples from $p(\mathbf{x})$.

A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in @def-explainability. If we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals, this is an expected finding, because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$. To generate samples, @grathwohl2020your rely on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization. This is where we depart from their methodology: instead of generating samples through SGLD, we propose using counterfactual explainers to generate counterfactuals for observed training samples. Specifically, we have

@@

Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while at same time increasing the energy of counterfactuals (negative samples) [@du2019implicit]. As the generated counterfactuals get more plausible (@def-explainability) over the cause of training, these two opposing effects gradually balance each out [@lippe2024uvadlc].

@@

Since actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to align model representations with user requirements. Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes through @eq-div. Here it is crucial that we avoid penalizing implausibility that arises due to mutability constraints. For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0$ whenever perturbing $\mathbf{x}^\prime[d]$ in the direction of $\mathbf{x}[d]$ would violate mutability constraints. Specifically, we set $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ if

@@

From a Bayesian perspective, setting $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x})$ with respect to feature $d$. Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments. Under certain conditions, this results holds theoretically[For the proof, see the supplementary appendix.]:

@@

Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ (i.e. *multinomial logistic regression*) where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$. If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty (@eq-div) will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is mutable and discriminative.

@@

Suppose we are interested in predicting the likelihood that loan applicants default on their credit. We have access to historical data on previous loan takers comprised of a binary outcome variable ($y\in\{1=\text{default},2=\text{no default}\}$) two input features: 1) the subjects' *age*, which we define as immutable, and 2) the subjects' existing level of *debt*, which we define as mutable.

@@

In panel (c), we have used conventional training again, this time imposing the mutability constraint on *age* at test time. Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals, in particular very young applicants. By comparison, counterfactual paths are shorter on average in panel (d), where we have used counterfactual training and protected immutable features as described in @sec-constraints. In particular, we observe that due to the classifier's lower sensitivity to *age*, recourse recommendations with respect to *debt* are much more homogenous, in that they do not disproportionately punish younger individuals. The counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to @def-explainability.

>>>>

Counterfactual training combines ideas from adversarial training, energy-based modelling and counterfactuals explanations with the explicit objective of aligning representations with plausible explanations that comply with user requirements. In the context of CEs, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful]. Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals affected by opaque models in practice. This is because for recourse recommendations to be **actionable**, they need to not only result in plausible counterfactuals but also be attainable. A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they were 20 years older. Ignoring all other features, this would comply with the definition of plausibility if 40-year-old individuals were in fact more credit-worthy on average than young adults. But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a (directly) mutable feature. For our intents and purposes, counterfactual training aims to improve model explainability by aligning models with counterfactuals that meet both desiderata, plausibility and actionability. Formally, we define explainability as follows:

@@

1. (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
2. (Actionability) Permutations $\Delta$ are subject to some actionability constraints.

@@

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual explanation generated through gradient descent over $T$ iterations as initially proposed by @wachter2017counterfactual. For our purposes, we let $T$ vary and consider the counterfactual search as converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.^[For detailed background information on gradient-based counterfactual search and convergence see @sec-app-ce.]

To train models with high explainability as defined in @def-explainability, we propose to leverage counterfactuals in the following objective:

@@

where $\text{yloss}(\cdot)$ is any conventional classification loss that induces discriminative performance (e.g., cross-entropy). The two additional components in @eq-obj are explained in more detail below. For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between counterfactuals $\mathbf{x}_T^\prime$ and observed samples $x$ and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. The tradeoff between the different components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$.

@@

@grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on inputs and generate inputs. They show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$. The first factor can be optimized using conventional cross-entropy as in @eq-obj. To optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimize the contrastive divergence between samples drawn from $p_\theta(\mathbf{x})$ and training observations, i.e., samples from $p(\mathbf{x})$.

A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in @def-explainability. If we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals, this is an expected finding, because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$. To generate samples, @grathwohl2020your rely on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization. This is where we depart from their methodology: instead of generating samples through SGLD, we propose using counterfactual explainers to generate counterfactuals for observed training samples. Specifically, we have:

@@

Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while at same time increasing the energy of counterfactuals (negative samples) [@du2019implicit]. As the generated counterfactuals get more plausible (@def-explainability) over the course of training, these two opposing effects gradually balance each out [@lippe2024uvadlc].

@@

Since actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to align model representations with user requirements. Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes through @eq-div. Here it is crucial that we avoid penalizing implausibility that arises due to mutability constraints. For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0$ whenever perturbing $\mathbf{x}^\prime[d]$ in the direction of $\mathbf{x}[d]$ would violate mutability constraints. Specifically, we set $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ if:

@@

From a Bayesian perspective, setting $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x})$ with respect to feature $d$. Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments. Under certain conditions, this results holds theoretically:^[For the proof, see the supplementary appendix.]

@@

Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ (i.e., *multinomial logistic regression*) where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$. If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty (@eq-div) will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.

@@

Suppose we are interested in predicting the likelihood that loan applicants default on their credit. We have access to historical data on previous loan takers comprised of a binary outcome variable ($y\in\{1=\text{default},2=\text{no default}\}$) two input features: (1) the subjects' *age*, which we define as immutable, and (2) the subjects' existing level of *debt*, which we define as mutable.

@@

In panel (c), we have used conventional training again, this time imposing the mutability constraint on *age* at test time. Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals (very young applicants). By comparison, counterfactual paths are shorter on average in panel (d), where we have used counterfactual training and protected immutable features as described in @sec-constraints. In particular, we observe that due to the classifier's lower sensitivity to *age*, recourse recommendations with respect to *debt* are much more homogenous, in that they do not disproportionately punish younger individuals. The counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to @def-explainability.

### 2025-03-10 paper/sections/other/compute.qmd

We performed our experiments on a high-performance cluster. Details about the cluster will be disclosed upon publication to avoid revealing information that might interfere with the double-blind review process. Since our experiments involve highly parallel tasks and rather small models by today's standard, we have relied on distributed computing across multiple central processing units (CPU). Graphical processing units (GPU) were not required. For larger grid searches we used up to but typically less than 100 CPUs.

>>>>

We performed our experiments on a high-performance cluster. Details about the cluster will be disclosed upon publication to avoid revealing information that might interfere with the double-blind review process. Since our experiments involve highly parallel tasks and rather small models by today's standard, we have relied on distributed computing across multiple central processing units (CPU). Graphical processing units (GPU) were not required. Model training for the largest grid searches with 270 unique parameter combinations was parallelized across 34 CPUs with 2GB memory each. The time to completion varied by dataset for reasons discussed in @sec-discussion: 0h49m (*Moons*), 1h4m (*Linearly Separable*), 1h49m (*Circles*), 3h52m (*Overlapping*). Model evaluations for large grid searches were parallelized across 20 CPUs with 3GB memory each. Evaluations for all data sets took less than one hour (<1h) to complete.
